# Reproducible Research: Peer Assessment 1

This document represents the solution to the Peer Assignment 1 project for the 
Reproducible Research class.

The solution consists of a sequence of basic data processing tasks. The detailed 
description of the data set and the requirements for each processing step are
presented in the [instructions](https://github.com/onlineclass/RepData_PeerAssessment1/blob/master/original.repo.README.md) document.

In the implementation of this solution, I chose to use only the base plotting system


## Loading and preprocessing the data

Load the data and store it in the data frame object **activity**

```{r}
activity <- read.csv("activity.csv", stringsAsFactors = F)
```


## What is mean total number of steps taken per day?

Use **tapply** function to create an array of total number of steps for each day and 
then plot the daily activity histogram

```{r}
activity.daily <- tapply(activity$steps, activity$date, sum, simplify = T)
hist(activity.daily, main = "Activity (NA values removed)", xlab = "Daily total steps", 
     ylab = "Frequency [Days]", col = "red")
```

The mean and median values of the daily totla steps are simply computed using the **mean** and **median** functions applied to the **activity.daily** (the NA values will be removed - **na.rm = T**)

```{r}
activity.daily.median <- median(activity.daily, na.rm = T)
activity.daily.mean <- mean(activity.daily, na.rm = T)
```

```{r}
activity.daily.median
```

```{r}
activity.daily.mean
```

## What is the average daily activity pattern?

The following steps will create the timeseries plot of the average number of 
steps in each 5-minute interval: 
* remove all **NA** rows from the original data and store the filtered set in **activity.interval**.
* create a new data frame - **activity.interval.avg** - using the **tapply** with the **mean** function
* plot **mean_steps** column as a function of **interval**

```{r}
activity.interval <- activity[!is.na(activity$steps),]

activity.interval.avg <- data.frame(
    interval = as.numeric(levels(as.factor(activity.interval$interval))), 
    mean_steps = tapply(activity.interval$steps, activity.interval$interval, mean))

plot(activity.interval.avg$interval, activity.interval.avg$mean_steps, type="l", 
     main = "5-minute interval - average steps", xlab = "interval", 
     ylab = "average steps")
```


The maximum activity interval will be determined with the following statement:

```{r}
maximum.activity.interval <- activity.interval.avg[
    activity.interval.avg$mean_steps == max(activity.interval.avg$mean_steps),1]

maximum.activity.interval
```





## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
